---
title: Genetic scores for diet response prediction
output: html_document
---

```{r setup, include=F}
knitr::opts_chunk$set(echo=F, cache.path="../cache/test/",
                      fig.path="../output/figures/test/")
suppressMessages(silent <- lapply(
  c("knitr", "tidyverse", "cowplot", "doParallel", "kableExtra",
    "glmnet", "broom"), library, character.only=T))
```

```{r helpers}
INT <- function(x) qnorm((rank(x, na.last="keep") - 0.5) / sum(!is.na(x)))

winsorize <- function(x, thresh) {
  bounds <- quantile(x, c(thresh, 1 - thresh), na.rm=T)
  case_when(x < bounds[1] ~ bounds[1],
            x > bounds[2] ~ bounds[2],
            TRUE ~ x)
} 

make_qqplot <- function(p_vec, plotTitle="Title") {
  p_vec <- p_vec[!is.na(p_vec)]
  qqplot(-log10(1:length(p_vec) / length(p_vec)), -log10(p_vec), pch=".", 
         main=plotTitle, xlab="Expected (-logP)", ylab="Observed (-logP)")
  abline(0, 1, col="red")
}
```

```{r load-metadata, message=F}
raw_phenos_whi <- read_csv("../data/processed/metadata_whi.csv", 
                           col_types=cols(sex="c", lipid_med="l", dm_trial="l",
                                          dm_intervention="l"))

whi_sample_to_subject <- read_tsv("../data/raw/whi/sample_info.txt", skip=15) %>%
  rename(sampleID=SampleID, subjID=SubjectID) %>%
  select(sampleID, subjID)
```

```{r clean-metadata, message=F}
clean_phenos_whi <- raw_phenos_whi %>%
  mutate(ht_med=as.logical(ht_med),
         ldl=ifelse(lipid_med, ldl / 0.75, ldl),
         glu=ifelse(dm_med, glu / 0.75, glu),
         sbp=ifelse(ht_med, sbp + 15, sbp),
         logf2c=log((sfa + mufa + pufa) / carb),
         logf2c=winsorize(logf2c, 0.005))

process_metadata <- function(clean_phenos, dv, rf, 
                             races=c("white", "black", "hispanic", "asian")) {
  phenos <- clean_phenos %>%
    filter(!is.na(.data[[dv]]),
           !is.na(.data[[rf]]),
           race %in% races)
  
  phenos_baseline <- phenos %>%
    filter(visit_year == 0)
  
  phenos_long <- phenos %>%
    mutate(arm=ifelse(dm_intervention, "intervention", "control")) %>%
    group_by(subjID, age, race, sex, dm_trial, arm) %>%
    filter(n() > 1) %>%
    arrange(visit_year) %>%
    summarise(baseline_bmi=bmi[1],
              baseline_rf=.data[[rf]][1],
              baseline_dv=.data[[dv]][1],
              delta_bmi=bmi[n()] - bmi[1],
              delta_rf=.data[[rf]][n()] - .data[[rf]][1],
              delta_dv=.data[[dv]][n()] - .data[[dv]][1]) %>%
    ungroup()
  
  list(baseline=phenos_baseline, long=phenos_long)
}
```

```{r add-genotypes}
add_genotypes <- function(phenos_list, genotype_path) {
  whi_genotypes <- readRDS(genotype_path) %>%
    rename(sampleID=IID) %>%
    select(-FID, -PAT, -MAT, -SEX, -PHENOTYPE) %>%
    mutate_all(Hmisc::impute, median)
  
  baseline_data <- phenos_list$baseline %>%
    inner_join(whi_sample_to_subject, by="subjID") %>%
    inner_join(whi_genotypes, by="sampleID")
  
  long_data <- phenos_list$long %>%
    inner_join(whi_sample_to_subject, by="subjID") %>%
    inner_join(whi_genotypes, by="sampleID")
  
  list(baseline=baseline_data, long=long_data)
}
```

```{r specific-experiments, cache=1, warning=F}
cs_trial_compare <- function(data_list, dv, rf) {
  cs_data <- filter(data_list$baseline, dm_trial == F)
  X <- as.matrix(select(cs_data, contains("rs")))
  y <- INT(scale(cs_data[[dv]]) * scale(log(cs_data[[rf]])))
  cs_nondm <- foreach(x=iter(X, by="col"), .combine=rbind,
                      .packages=c("broom")) %dopar%
    tidy(lm(y ~ x))[2, ]
  cs_nondm$snp <- colnames(X)
  
  trial_data <- filter(data_list$long, 
                       dm_trial == T, arm == "intervention")
  X <- as.matrix(select(trial_data, contains("rs")))
  y <- INT(trial_data$delta_rf)
  direct_effect_trial <- foreach(x=iter(X, by="col"), .combine=rbind,
                                 .packages=c("broom")) %dopar%
    tidy(lm(y ~ x))[2, ]
  direct_effect_trial$snp <- colnames(X)
  
  cor_test <- cor.test(cs_nondm$estimate, direct_effect_trial$estimate,
                       method="spearman")
  list(cs_nondm=cs_nondm, direct_effect_trial=direct_effect_trial,
       cor_test=cor_test, corr=cor_test$estimate)
}

full_white_test <- function(dv, rf, genopath) {
  a <- process_metadata(clean_phenos_whi, dv, rf, "white")
  b <- add_genotypes(a, genopath)
  cs_trial_compare(b, dv, rf)
}

rf_to_path <- list(ldl="../data/processed/whi_ldl_suggestive.rds",
                   chol="../data/processed/whi_ldl_suggestive.rds",
                   glu="../data/processed/whi_glu_suggestive.rds",
                   bmi="../data/processed/whi_bmi_genomewide16.rds",
                   sbp="../data/processed/whi_bp_suggestive.rds")

p_thresh_test <- function(test_res, de_res, p_thresh=0.2) {
  shared_snps <- intersect(test_res$snp, de_res$snp)
  cor_test_df <- data.frame(cs=test_res$estimate[match(shared_snps, test_res$snp)],
                            de=de_res$estimate[match(shared_snps, de_res$snp)])
  tidy(cor.test(~ cs + de, data=cor_test_df, method="spearman",
                subset=test_res$p.value < p_thresh))$estimate
}
```

The "sanity check/is there any hope" strategy:

* Restrict to European ancestry only
* For each risk factor tested, subset to only genome-wide suggestive (p<1e-5) SNPs from GWAS for that risk factor
* Test those SNPs for association with 1) our "interaction product method" (diet variable = SFA or fat:carb ratio), and 2) the change in that risk factor over the first three years of the WHI diet trial
* Test the Spearman correlation between SNP coefficient estimates for the two regression sets above. A negative correlation is "good", in the sense that predicted responders (positive product values) have a greater *decrease* in risk factor during the trial.

```{r show-corrs, warning=F, eval=F}
# mygrid$spearman_corr <- map_dbl(mycorrs, "estimate")
mygrid$spearman_corr <- map_dbl(mycorrs, "corr")
ggplot(mygrid, aes(x=dv, y=rf, fill=spearman_corr)) +
  geom_tile() +
  scale_fill_gradient2() +
  labs(subtitle="Spearman correlation of all betas")

p_thresh_test <- function(test_res, de_res, p_thresh=0.2) {
  shared_snps <- intersect(test_res$snp, de_res$snp)
  cor_test_df <- data.frame(cs=test_res$estimate[match(shared_snps, test_res$snp)],
                            de=de_res$estimate[match(shared_snps, de_res$snp)])
  tidy(cor.test(~ cs + de, data=cor_test_df, method="spearman",
                subset=test_res$p.value < p_thresh))$estimate
}

mygrid$spearman_corr_with_thresh <- map_dbl(
  mycorrs, function(x) p_thresh_test(x$cs_nondm, x$direct_effect_trial))
ggplot(mygrid, aes(x=dv, y=rf, fill=spearman_corr_with_thresh)) +
  geom_tile() +
  scale_fill_gradient2() +
  labs(subtitle="Spearman correlation of betas for SNPs w/ C-S p<0.2")
```

## This time, testing SFA -> SBP

### What if this is done in different ancestries (non-European)?

```{r ancestries, cache=1, warning=F}
chosen_dv <- "sfa"
chosen_rf <- "sbp"

white_only <- process_metadata(
  clean_phenos_whi, chosen_dv, chosen_rf, "white") %>%
  add_genotypes(rf_to_path[[chosen_rf]])
black_only <- process_metadata(
  clean_phenos_whi, chosen_dv, chosen_rf, "black") %>%
  add_genotypes(rf_to_path[[chosen_rf]])
both_ancestries <- process_metadata(
  clean_phenos_whi, chosen_dv, chosen_rf, c("white", "black")) %>% 
  add_genotypes(rf_to_path[[chosen_rf]])

cl <- makePSOCKcluster(8)
registerDoParallel(cl)

white_corr <- cs_trial_compare(white_only, "sfa", "sbp")
black_corr <- cs_trial_compare(black_only, "sfa", "sbp")
both_corr <- cs_trial_compare(both_ancestries, "sfa", "sbp")

stopCluster(cl)
```

```{r show-ancestries, warning=F}
ancestry_res <- data.frame(
  Race=c("White", "Black", "Both"),
  Spearman_corr=c(p_thresh_test(white_corr$cs_nondm, white_corr$direct_effect_trial),
                  p_thresh_test(black_corr$cs_nondm, black_corr$direct_effect_trial),
                  p_thresh_test(both_corr$cs_nondm, both_corr$direct_effect_trial)))

kable(ancestry_res) %>% kable_styling()
```

It seems that this strong negative correlation is preserved only in whites (though sample size may play a role here).

### Negative control: what about using a non-matching set of SNPs?

```{r test-random-snps, cache=1, warning=F}
white_only_with_glu <- process_metadata(
  clean_phenos_whi, chosen_dv, chosen_rf, "white") %>% 
  add_genotypes(rf_to_path[["glu"]])
white_only_with_bmi <- process_metadata(
  clean_phenos_whi, chosen_dv, chosen_rf, "white") %>% 
  add_genotypes(rf_to_path[["bmi"]])

cl <- makePSOCKcluster(8)
registerDoParallel(cl)
mismatch_res <- cs_trial_compare(white_only_with_glu, chosen_dv, chosen_rf)
mismatch_res2 <- cs_trial_compare(white_only_with_bmi, chosen_dv, chosen_rf)
stopCluster(cl)
```

```{r show-test-random-snps, warning=F}
# print("SFA -> SBP beta correlation using Glu-related SNPs:")
# p_thresh_test(mismatch_res$cs_nondm, mismatch_res$direct_effect_trial)
# print("SFA -> SBP beta correlation using BMI-related SNPs:")
# p_thresh_test(mismatch_res2$cs_nondm, mismatch_res2$direct_effect_trial)

random_snps_res_df <- data.frame(
  `SNP set`=c("Glucose", "BMI"),
  `Spearman_corr`=c(p_thresh_test(mismatch_res$cs_nondm, mismatch_res$direct_effect_trial),
                    p_thresh_test(mismatch_res2$cs_nondm, mismatch_res2$direct_effect_trial))
)

kable(random_snps_res_df) %>% kable_styling()
```

### Negative "control": use SBP main effect estimates instead of interaction product

```{r test-main-effect, cache=1}
cl <- makePSOCKcluster(8)
registerDoParallel(cl)

cs_data <- filter(white_only$baseline, dm_trial == F)
X <- as.matrix(select(cs_data, contains("rs")))
y <- INT(log(cs_data[[chosen_rf]]))
cs_nondm_main <- foreach(x=iter(X, by="col"), .combine=rbind,
                    .packages=c("broom")) %dopar%
  tidy(lm(y ~ x))[2, ]
cs_nondm_main$snp <- colnames(X)

stopCluster(cl)
```

```{r show-test-main-effect, warning=F}
# tidy(cor.test(cs_nondm_main$estimate, 
#               white_corr$direct_effect_trial$estimate,
#               method="spearman"))
p_thresh_test(cs_nondm_main, white_corr$direct_effect_trial)
```

### Prediction algorithm performance (CS -> trial)

```{r prediction, cache=1}
cs_data <- filter(white_only$baseline, dm_trial == F)
use_snps <- white_corr$cs_nondm$snp[white_corr$cs_nondm$p.value < 0.01]
y <- INT(scale(cs_data[[chosen_dv]]) * scale(log(cs_data[[chosen_rf]])))
glm_data <- select(cs_data, use_snps)
glm_fit <- glm(y ~ ., data=glm_data)
# X <- as.matrix(select(cs_data, contains("rs")))
# y <- INT(scale(cs_data[[chosen_dv]]) * scale(log(cs_data[[chosen_rf]])))
# use_snps <- colnames(X) ######[ldl_cs_nondm$p.value < 0.1]
# cl <- makePSOCKcluster(4)
# registerDoParallel(cl)
# glm_fit <- cv.glmnet(X[, use_snps], y, alpha=0.5, nfolds=3, parallel=T)
# stopCluster(cl)
# print(sum(coef(glm_fit, s="lambda.min") != 0))

trial_data <- filter(white_only$long, dm_trial == T)
y_pred <- as.vector(predict(glm_fit, trial_data))
tidy(lm(delta_rf ~ y_pred, data=trial_data, 
        subset=trial_data$arm == "intervention"))

# trial_data <- filter(white_only$long, dm_trial == T)
# X_test <- as.matrix(select(trial_data, contains("rs")))
# y_pred <- as.vector(predict(glm_fit, X_test[, use_snps], s="lambda.min"))
# tidy(lm(delta_rf ~ y_pred, data=trial_data, 
#         subset=trial_data$arm == "intervention"))
```

```{r prediction-alpha-zero, cache=1}
cs_data <- filter(white_only$baseline, dm_trial == F)
X <- as.matrix(select(cs_data, contains("rs")))
y <- INT(scale(cs_data[[chosen_dv]]) * scale(log(cs_data[[chosen_rf]])))
use_snps <- colnames(X) ######[ldl_cs_nondm$p.value < 0.1]
cl <- makePSOCKcluster(4)
registerDoParallel(cl)
glm_fit <- cv.glmnet(X[, use_snps], y, alpha=0, nfolds=3, parallel=T)
stopCluster(cl)
print(sum(coef(glm_fit, s="lambda.min") != 0))

trial_data <- filter(white_only$long, dm_trial == T)
X_test <- as.matrix(select(trial_data, contains("rs")))
y_pred <- as.vector(predict(glm_fit, X_test[, use_snps], s="lambda.min"))
tidy(lm(delta_rf ~ y_pred, data=trial_data,
        subset=trial_data$arm == "intervention"))
```

### Refining: do any alternative statistical approaches improve the beta correlation?

```{r variations, cache=1}
cl <- makePSOCKcluster(4)
registerDoParallel(cl)

cs_data <- filter(white_only$baseline, dm_trial == F)
X <- as.matrix(select(cs_data, contains("rs")))

y <- scale(cs_data[[chosen_dv]]) * scale(log(cs_data[[chosen_rf]]))
cs_nondm_noINT <- foreach(x=iter(X, by="col"), .combine=rbind,
                          .packages=c("broom")) %dopar%
  tidy(lm(y ~ x))[2, ]
cs_nondm_noINT$snp <- colnames(X)

y <- log(cs_data[[chosen_rf]])
cs_nondm_interaction <- foreach(x=iter(X, by="col"), .combine=rbind,
                                .packages=c("broom")) %dopar%
  tidy(lm(y ~ x * cs_data[[chosen_dv]], data=cs_data))[4, ]
cs_nondm_interaction$snp <- colnames(X)

stopCluster(cl)
```

```{r show-variations, warning=F}
# cor.test(cs_nondm_noINT$estimate, white_corr$direct_effect_trial$estimate, method="spearman")
# cor.test(cs_nondm_interaction$estimate, white_corr$direct_effect_trial$estimate, method="spearman")
variations_res_df <- data.frame(
  Variant=c("No inverse-normal transform of phenotype",
            "Interaction rather than 'product method'"),
  p=c(p_thresh_test(cs_nondm_noINT, white_corr$direct_effect_trial),
      p_thresh_test(cs_nondm_interaction, white_corr$direct_effect_trial)))

kable(variations_res_df) %>% kable_styling()
```

### Delta method

Right now, with logF2C -> SBP

```{r delta-method-comparison, cache=1}
cl <- makePSOCKcluster(8)
registerDoParallel(cl)

f2c_sbp_data <- process_metadata(
  clean_phenos_whi, "logf2c", "sbp", "white") %>%
  add_genotypes(rf_to_path$sbp)
long_nondm_data <- filter(f2c_sbp_data$long, dm_trial == F)
X <- as.matrix(select(long_nondm_data, contains("rs")))
y <- INT(scale(long_nondm_data$delta_dv) * scale(long_nondm_data$delta_rf))
long_nondm <- foreach(x=iter(X, by="col"), .combine=rbind,
                    .packages=c("broom")) %dopar%
  tidy(lm(y ~ x))[2, ]
long_nondm$snp <- colnames(X)

stopCluster(cl)
```

```{r show-delta_method_comparison}
delta_method_corr <- p_thresh_test(long_nondm, white_corr$direct_effect_trial)

paste("Spearman corr. using delta method (change in DV x change in RF):", 
      delta_method_corr)
```

### Other cohorts -- also correlated?

```{r fhs-comparison, cache=1, message=F, eval=F}
fhs_meta <- read_csv("../data/processed/metadata_fhs.csv") %>%
  mutate(sbp=ifelse(ht_med, sbp + 15, sbp)) %>%
  filter(!is.na(sfa), !is.na(sbp))
fhs_genos <- readRDS("../data/processed/fhs_ldl_suggestive.rds") %>%
  rename(subjID=IID) %>%
  select(-FID, -PAT, -MAT, -SEX, -PHENOTYPE) %>%
  mutate_all(Hmisc::impute, median)
fhs_data <- inner_join(fhs_meta, fhs_genos, by="subjID")

cl <- makePSOCKcluster(8)
registerDoParallel(cl)

fhs_X <- as.matrix(select(fhs_data, contains("rs")))
fhs_y <- INT(scale(fhs_data[[chosen_dv]]) * scale(log(fhs_data[[chosen_rf]])))
cs_fhs <- foreach(x=iter(fhs_X, by="col"), .combine=rbind,
                    .packages=c("broom")) %dopar%
  tidy(lm(fhs_y ~ x))[2, ]
cs_fhs$snp <- colnames(fhs_X)

fhs_f_data <- filter(fhs_data, sex == "F")
fhs_f_X <- as.matrix(select(fhs_f_data, contains("rs")))
fhs_f_y <- INT(scale(fhs_f_data[[chosen_dv]]) * scale(log(fhs_f_data[[chosen_rf]])))
cs_fhs_f <- foreach(x=iter(fhs_f_X, by="col"), .combine=rbind,
                    .packages=c("broom")) %dopar%
  tidy(lm(fhs_f_y ~ x))[2, ]
cs_fhs_f$snp <- colnames(fhs_f_X)

stopCluster(cl)
```

```{r show-fhs-comparison, warning=F, eval=F}
# whi_fhs_shared_snps <- intersect(white_corr$direct_effect_trial$snp,
#                                  cs_fhs$snp)
# whi_de_res <- white_corr$direct_effect_trial
# print("FHS-CS -> WHI DM effect correlation")
# tidy(cor.test(whi_de_res$estimate[match(whi_fhs_shared_snps, whi_de_res$snp)],
#               cs_fhs$estimate[match(whi_fhs_shared_snps, cs_fhs$snp)],
#               method="spearman"))
# print("FHS-CS -> WHI DM effect correlation (females only)")
# tidy(cor.test(whi_de_res$estimate[match(whi_fhs_shared_snps, whi_de_res$snp)],
#               cs_fhs_f$estimate[match(whi_fhs_shared_snps, cs_fhs_f$snp)],
#               method="spearman"))
print("FHS-CS -> WHI DM effect correlation")
p_thresh_test(cs_fhs, white_corr$direct_effect_trial)
print("FHS-CS -> WHI DM effect correlation (females only)")
p_thresh_test(cs_fhs_f, white_corr$direct_effect_trial)
```

```{r mesa-comparison, cache=1, message=F, eval=F}
mesa_meta <- read_csv("../data/processed/metadata_mesa.csv") %>%
  mutate(ldl=ifelse(statin, ldl / 0.75, ldl)) %>%
  filter(!is.na(sfa), !is.na(ldl))
mesa_white_genos <- readRDS("../data/processed/mesa_white_ldl_suggestive.rds") %>%
  rename(subjID=IID) %>%
  select(-FID, -PAT, -MAT, -SEX, -PHENOTYPE) %>%
  mutate_all(Hmisc::impute, median)
mesa_white_data <- inner_join(mesa_meta, mesa_white_genos, by="subjID")

cl <- makePSOCKcluster(8)
registerDoParallel(cl)

mesa_white_X <- as.matrix(select(mesa_white_data, contains("rs")))
mesa_white_y <- INT(scale(mesa_white_data[[chosen_dv]]) * 
                      scale(log(mesa_white_data[[chosen_rf]])))
cs_mesa_white <- foreach(x=iter(mesa_white_X, by="col"), .combine=rbind,
                    .packages=c("broom")) %dopar%
  tidy(lm(mesa_white_y ~ x))[2, ]
cs_mesa_white$snp <- colnames(mesa_white_X)

mesa_white_f_data <- filter(mesa_white_data, sex == "F")
mesa_white_f_X <- as.matrix(select(mesa_white_f_data, contains("rs")))
mesa_white_f_y <- INT(scale(mesa_white_f_data[[chosen_dv]]) * 
                      scale(log(mesa_white_f_data[[chosen_rf]])))
cs_mesa_white_f <- foreach(x=iter(mesa_white_f_X, by="col"), .combine=rbind,
                    .packages=c("broom")) %dopar%
  tidy(lm(mesa_white_f_y ~ x))[2, ]
cs_mesa_white_f$snp <- colnames(mesa_white_f_X)

stopCluster(cl)
```

```{r show-mesa-comparison, warning=F, eval=F}
# whi_mesa_white_shared_snps <- intersect(white_corr$direct_effect_trial$snp,
#                                         cs_mesa_white$snp)
# whi_de_res <- white_corr$direct_effect_trial
# print("MESA-CS -> WHI DM effect correlation")
# tidy(cor.test(whi_de_res$estimate[match(whi_mesa_white_shared_snps, whi_de_res$snp)],
#               cs_mesa_white$estimate[match(whi_mesa_white_shared_snps, cs_mesa_white$snp)],
#               method="spearman"))
# print("MESA-CS -> WHI DM effect correlation (females only)")
# tidy(cor.test(whi_de_res$estimate[match(whi_mesa_white_shared_snps, whi_de_res$snp)],
#               cs_mesa_white_f$estimate[match(whi_mesa_white_shared_snps, cs_mesa_white_f$snp)],
#               method="spearman"))
print("MESA-CS -> WHI DM effect correlation")
p_thresh_test(cs_mesa_white, white_corr$direct_effect_trial)
print("MESA-CS -> WHI DM effect correlation (females only")
p_thresh_test(cs_mesa_white_f, white_corr$direct_effect_trial)
```